---
title: "Notes"
author: "Ben Welman"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cerulean
    highlight: kate
---

```{r setup, echo=FALSE}
suppressPackageStartupMessages(library(fpp3))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(kableExtra))
```

# 1. Getting started

A time series is a sequence of observations collected at time intervals. A time series object has a time component in it.

# 2. Time series graphics

## tsibble objects

A `tsibble` object, built on a tibble, has the following columns:

+ A column for the time component, called the **index**.
+ One or more columns for the values of the observations
+ Optional one or more columns for the characteristics of the subjects of the observations, together called the **key**

Just like a database: **Each row (observation) must be uniquely identified by index and key.**

### The index variable {-}

**Yearly data**

```{r}
y <- tsibble(
	Year = 2015:2019,
	Observation = c(123, 39, 78, 52, 110),
	index = Year
)
y
```

**Quarterly data**

```{r}
x <- tibble(Quarter = seq(as.Date("2020-01-01"), as.Date("2020-12-31"), by = "1 quarter"),
			Observation = c(48, 73, 64, 120))
y <- x %>%
	mutate(Quarter = yearquarter(Quarter)) %>%
	as_tsibble(index = Quarter)
y
```

**Monthly data**

```{r}
x <- tibble(Month = seq(as.Date("2020-01-01"), as.Date("2020-12-31"), by = "1 month"),
			Observation = c(106, 58, 92, 33, 23, 70, 119, 41, 79, 29, 73, 58))
y <- x %>%
	mutate(Month = yearmonth(Month)) %>%
	as_tsibble(index = Month)
y
```

**Weekly data**

```{r}
x <- tibble(Week = seq(as.Date("2020-01-01"), as.Date("2020-06-30"), by = "1 week"),
			Observation = sample(100:500, size = 26))
y <- x %>%
	mutate(Week = yearweek(Week)) %>%
	as_tsibble(index = Week)
y
```

**Daily data**

```{r}
x <- tibble(Day = seq(as.Date("2020-01-01"), as.Date("2020-01-31"), by = "1 day"),
			Observation = sample(500:600, size = 31))
y <- x %>%
	as_tsibble(index = Day)
y
```

**half-hourly data**

```{r}
vic_elec
```

**ts data > tsibble**

```{r}
x <- astsa::globtempl # a ts object
y <- as_tsibble(x)
y
```

**data frame > tsibble>


### The key variables {-}

Example: the ansett dataset is a weekly tsibble with data of the passenger numbers on Ansett airline flights.

```{r}
ansett
```

This is a `7407x4` tsibble object, 7407 rows and 4 columns. The columns are

+ `Week`: The time component. This is the index.
+ `Passengers`: Total air passengers travelling with Ansett. The obervation value.
+ `Airports`: The airports that passengers are travelling between (both directions)
+ `Class`: The class of the ticket.

The `[1W]` tells us that the time interval is 1 week.

The `Key: Airports, Class [30]` tells us that the keys are `Airports` and `Class` and that there are 30 unique combinations of `Airports` and `Class`.

```{r}
ansett %>% distinct(Airports)
```

10 different values for the variable `Airports`.

```{r}
ansett %>% distinct(Class)
```

3 different values for the variable `Class`.

So in fact there are 30 different time series.

You can get one of these 30 time series using `filter()`

```{r}
ansett %>% 
	filter(Airports == "ADL-PER" & Class == "Business") %>%
	autoplot()
```

### Conversions {-}

See

**ts data > tsibble**

```{r}
x <- astsa::globtempl # a ts object
y <- as_tsibble(x)
y
```

**data frame > tsibble>**

To coerce a data frame to tsibble, you need to declare key and index. See https://tsibble.tidyverts.org/

# Data sets

```{r echo=FALSE}
ds <- function(p) {
	vcdExtra::datasets(p) %>% 
	select(-class) %>%
	knitr::kable() %>% kable_styling()
}
```

**fpp3 package**

```{r echo=FALSE}
ds("fpp3")
```

**tsibble**

```{r echo=FALSE}
ds("tsibble")
```

**tsibbledata**

```{r echo=FALSE}
ds("tsibbledata")
```


# References

+ [Forecasting: Principles and Practice (3rd ed)](https://otexts.com/fpp3/)

+ [ETC3550Slides, Hyndman](https://github.com/robjhyndman/ETC3550Slides/tree/fable)

+ [tsibbledata reference](https://tsibbledata.tidyverts.org/reference/index.html)

